---
title: "HW1"
author: Sandra Villamar and Shobhit Dronamraju
output: pdf_document
date: "2023-01-19"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath(
  "C://Users/sandr/OneDrive/Documents/UCSD Masters/DSC 291 Stat Models")) 
```

```{r}
library(ggplot2)
```

1.  Write a function *confBand(x, y, conf=0.95)* taking in a predictor
    vector *x* and a response vector *y* and return a plot with (x, y)
    points, the least squares line, and the confidence band at level
    *conf*. Apply the function to hp and mpg from 04cars dataset.

```{r}
confBand <- function(x, y, conf=0.95){

  # compute confidence band lines
  fit = lm(y ~ x)
  pred_interval = predict(fit, data.frame(x), interval='confidence')
  lower = pred_interval[,2]
  upper = pred_interval[,3]
  
  # plot
  plot(x, y, pch=16)  # points
  abline(fit$coefficients[1], fit$coefficients[2], col='blue') # reg line
  points(x, lower, col='red', type='l')  # lower band
  points(x, upper, col='red', type='l')  # upper band
}
```

```{r}
load("data/04cars.rda")  # load cars dataset to "dat"
dat = dat[,c(13,15,16,18,19)]  # extract selected variables
dat = dat[complete.cases(dat),]  # extract complete cases
names(dat) = c("hp","mpg","wt","len","wd")  # abbreviate names
str(dat)

confBand(dat$hp, dat$mpg)  # apply function

```

2.  Let n = 100. Draw n i.i.d samples from *Unif(0, 1)* which stay fixed
    in what follows. Repeat the following experiment N=1000 times.

-   Generate $y_i = 1 + x_i + \epsilon_i$ with $\epsilon_i$ i.i.d. *N(0,
    0.2)*
-   Compute the 99% confidence band and record whether or not it
    contains the true line or not

Summarize the result of this numerical experiment by returning the
proportion of times (out of N) that the confidence band contained the
true line.

```{r}
n = 100
N = 1000
x = runif(n, 0, 1)
y_true = 1 + x
p = 1
count = 0

for(i in 1:N){
  
  # setting up formula
  e = rnorm(100, 0, sqrt(0.2))
  y = 1 + x + e

  # computing variables for confidence band
  fit = lm(y~x)
  pred = predict(fit, data.frame(x), se.fit=TRUE)
  y_hat = pred$fit
  se_hat = pred$se.fit
  
  # compute confidence band
  interval = sqrt((p + 1) * qf(0.99, p + 1, n - p - 1)) * se_hat
  lower = y_hat - interval
  upper = y_hat + interval
  
  # check if true line within band
  if(all(lower <= y_true) && all(y_true <= upper)){
    count = count + 1
  }
}
 
proportion = count/N
```

The experiment produced the proportion: `r proportion`

## Contributions:

We worked on all parts of this assignment together.
