---
title: "Lab2"
output: html_document
---



```{r include=FALSE}
library(alr4)
library(car)
library(MASS)
library(ggplot2)
library(ellipse)
```


# Hypothesis testing

Consider the Boston data and the linear model: $medv = \beta_0 + \beta_1*crim + \beta_2*nox + \beta_3*rm + \beta_4*age + \beta_5*dis + \epsilon$
```{r}
lm.medv = lm(medv ~ crim + nox + rm + age + dis,data = Boston) 
summary(lm.medv)
```

## Hypothesis testing of estimated slope
Test the hypothesis that the slope $\beta_1 = 0$, against the alternative that $\beta_1 < 0$.

```{r}
tval <- (coef(lm.medv)[2] - 0)/ sqrt(vcov(lm.medv)[2,2]) 
df <- dim(Boston)[1] - 2
data.frame(tval = tval, df=df, pval = 1 - pt(abs(tval), df))
```

The p-value is `r 1 - pt(abs(tval), df)` < 0.05. Therefore we reject the null hypothesis.

```{r}
confint(lm.medv, "crim",level = 0.95)
```

If we want to test the hypothesis that $\beta_3 = \beta_4 = 0$:

```{r}
lm.medv.h0 = lm(medv ~ crim + nox + dis,data = Boston) 
anova(lm.medv.h0, lm.medv)
```


The p-value is less than 0.05. We reject the null hypothesis, claiming at least one of the coefficient is not 0.

# Confidence region for regression coefficients 

```{r}
plot(ellipse(lm.medv,which = c(4,5),level = 0.95),type = 'l',col = 'darkred')
points(lm.medv$coefficients[4],lm.medv$coefficients[5],col = 'darkblue',pch = 18)
```


## Coverage probability of confidence intervals

The confidence level represents the long-run proportion of corresponding CIs that contain the true value of the parameter. For example, out of all intervals computed at the 95\% level, 95\% of them should contain the parameter's true value.

How do we validate this? Simulation

```{r}
B = 1000
beta0 = 1
beta1 = 3
n = 100 
CI = data.frame(lower = rep(NA,B),upper = rep(NA,B),contain_true_value = rep(NA,B))

for (i in 1:B) {
  x = rnorm(n)
  y = beta0 + beta1*x + 0.5*rnorm(n)
  lm.sim = lm(y~x)
  beta1_CI = confint(lm.sim, "x",level = 0.95)
  CI$lower[i] = beta1_CI[1]
  CI$upper[i] = beta1_CI[2]
  CI$contain_true_value[i] = (beta1_CI[1] <= beta1 & beta1_CI[2] >= beta1)
}
```

```{r}
cols = c('TRUE' = 'darkgreen','FALSE' = 'darkred')
size = c('TRUE' = 0.5,'FALSE' = 1)
ggplot(data = CI) + 
  geom_errorbar(aes(x = 1:B,y = beta1,ymin=lower, ymax=upper,color = contain_true_value,size = contain_true_value)) +
  geom_hline(yintercept = beta1) +
  scale_color_manual(values = cols) + 
  scale_size_manual(values = size) +
  xlab('Number of experiments')
```

Validate coverage probability:

```{r}
sum(CI$contain_true_value)/B
```

This estimated coverage probability will coverge to 0.95 as $B \to \infty$.