---
title: "Lab6"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(car)
library(alr4)
library(boot)
```

# Bootstrap for a median

## Using boot function
```{r message=FALSE, warning=FALSE}
attach(UN11)
# a function that will return the median
get.median <- function(data, indices) { median(data[indices]) }

set.seed(241)
b1 <- boot(UN11[,c('fertility')], get.median, R=1000) 
boot.ci(b1, type="all")
```

The function `boot.ci` returns confidence intervals at 95\% by default. Intervals can be computed in several ways for the bootstrap samples: `basic` uses the boostrap pivotal confidence interval; `norm` assumes the bootstraped values are approximately normal and independent and computes the interval using the bootstrap mean plus or minus 1.96 times the bootstrap standard error; `perc` uses the percentile bootstrap; `stud`  uses the studentized  bootstrap  confidence  interval.

## Resampling from scratch
```{r}
N <- length(UN11$fertility)
n.boot <- 1000
dat.boot <- rep(NA, n.boot) ## for saving results
set.seed(241)
for (i in 1:n.boot) {
  indices <- sample(seq(1,N), replace=T) 
  dat.boot[i] <- median(UN11$fertility[indices])
}
cat('Percentile bootstrap confidence interval\n')
print(round(quantile(dat.boot, c(0.025,0.975)),4))
cat('Basic bootstrap confidence interval\n')
print(round(2*median(UN11$fertility) - quantile(dat.boot, c(0.975,0.025)),4))
```


# Bootstrap for regression coefficients

```{r}
B = 1000
N = dim(UN11)[1]
level = 0.95

# fit the linear model
fit = lm(lifeExpF~fertility,data = UN11)
beta0 = fit$coefficients[1]
beta1 = fit$coefficients[2]
sebeta0=summary(fit)$coefficients[,2][1]
sebeta1=summary(fit)$coefficients[,2][2]

summary(fit)
plot(UN11$fertility,UN11$lifeExpF)
abline(fit,col = 'darkred',lwd = 3)
```

```{r}
beta0_boot = rep(NA,N)
beta1_boot = rep(NA,N)
t0_boot = rep(NA,N)
t1_boot = rep(NA,N)

set.seed(241)
for (i in 1:B){
  indices = sample(1:N,N,replace=TRUE)
  x_boot = UN11$fertility[indices]
  y_boot = UN11$lifeExpF[indices]
  fit_boot=lm(y_boot~x_boot)
  beta0_boot[i] = fit_boot$coefficients[1]
  beta1_boot[i] = fit_boot$coefficients[2]
  sebeta0_boot = summary(fit_boot)$coefficients[,2][1]
  sebeta1_boot = summary(fit_boot)$coefficients[,2][2]
  t0_boot[i]=(beta0_boot[i]-beta0)/(sebeta0_boot)
  t1_boot[i]=(beta1_boot[i]-beta1)/(sebeta1_boot)
}

```

Confidence interval for the intercept:
```{r}
boot_int = matrix(c(beta0 + quantile(t0_boot,c((1-level)/2,(1+level)/2))*sebeta0),ncol = 2)
colnames(boot_int) = c('2.5 %','97.5 %')
cat('Parametric confidence interval\n')
confint(fit)[1,]
cat('studentized  bootstrap  confidence  interval\n')
boot_int
```

Confidence interval for the slope:
```{r}
boot_slp = matrix(c(beta1 + quantile(t1_boot,c((1-level)/2,(1+level)/2))*sebeta1),ncol = 2)
colnames(boot_slp) = c('2.5 %','97.5 %')
cat('Parametric confidence interval\n')
confint(fit)[2,]
cat('Studentized  bootstrap  confidence  interval\n')
boot_slp
```