---
title: "Lab7"
output: html_document
---

```{r include=FALSE}
library(car)
library(alr4)
library(leaps)
library(MASS)
library(caret)
```

## Model selection
```{r}
data("mantel")
mantel
m0 <- lm(Y ~ 1, data=mantel) 
step(m0,scope= ~ X1 + X2 + X3, direction="forward")
```

Starting with the mean function with no predictors, at the first step we consider adding the one term that makes AIC as small as possible, which is $X_3$. At the second step, we consider adding another term after $X_3$ if it further reduces AIC; in this problem adding either $X_1$ or $X_2$ actually increases AIC, so we would select $\{X_3\}$.

```{r}
m1 <- lm(Y ~ X1 + X2 + X3, data=mantel)
step(m1,scope=~1, direction="backward")
```

It appears that the backward elimination algorithm selects to remove only $X_3$. Using backward elimination, we select $\{X_1 , X_2 \}$. These two computational algorithms give different answers. We would certainly prefer the choice $\{X_1 , X_2 \}$ from backward elimination because it gives an exact fit.


## The Boston dataset

### Forward selection
```{r}
dat = Boston[, -c(4, 9)]
m0 = lm(medv ~ 1, data=dat) 
m1 = lm(medv ~ ., data=dat) 
step(m0,scope = formula(m1), direction="forward")
```
The forward selection alpgorithm selects $\{lstat , rm, ptratio, dis, nox, black, zn, crim \}$. AIC of the model selected by forward selection alpgorithm is 1614.5.

### Backward selection 
```{r}
step(m1, direction="backward")

```

The backward selection alpgorithm selects $\{crim, zn, nox, rm, dis, ptratio, black, lstat\}$. This is the same as the model selected by the forward selection alpgorithm.


### Best subset

```{r}
X = dat[,!names(dat) %in% c('medv')]
L = leaps(X, dat$medv)
ind = which.min(L$Cp)
plot(L$size,L$Cp)
points(L$size[ind],L$Cp[ind],col = 'darkred',pch = 19)
points(aggregate(L$Cp, by = list(L$size), min),lwd = 2,col = 'darkred',type = 'b',pch = 19)
```

Best model selected:

```{r}
names(X)[L$which[ind,]] 
```

The best subset selection alpgorithm selects $\{crim, zn, nox, rm, dis, ptratio, black, lstat\}$. This is the same as the model selected by the two greedy model selection algorithms.


### Prediction error estimated by k-fold cross-validation

Consider the model `medv ~ crim + zn + nox + rm + dis + ptratio + black + lstat`. We aim to use k-fold cross validate to estimate the RMSE of this model.

Split the dataset into blocks/folds
```{r}
set.seed(241)
k = 5
n = nrow(dat)
dat_cv = dat[sample(n),]
folds <- cut(seq(1,n),breaks=k,labels=FALSE)
folds
```

The model is fitted on $K − 1$ blocks and tested on the remaining block. Each block plays the role of validation data in turn. This results in $K$ different estimates for the prediction error. Their average is the cross-validation estimate. 

```{r}

cv_err = rep(NA, k)

for (i in 1:k) {
  dat_train = dat_cv[folds != i,]
  dat_val = dat_cv[folds == i,]
  
  m_train = lm(medv ~ crim + zn + nox + rm + dis + ptratio + black + lstat, data = dat_train)
  pred_val = predict(m_train, newdata = dat_val)
  res_val = dat_val$medv -pred_val
  cv_err[i] = sqrt(mean(res_val^2))
}

mean(cv_err)

```
 The estimated RMSE is `r mean(cv_err)`


Cross-validation using functions in the package `caret`:

```{r}
set.seed(241)
train_control <- trainControl(method = "cv",number = k)
m_cv <- train(medv ~ crim + zn + nox + rm + dis + ptratio + black + lstat, data = dat_cv,
               method = "lm",
               trControl = train_control)
m_cv
```

If we use the `caret` package, The estimated RMSE is `r m_cv$results['RMSE']`.